---
title: "Figure: CV methods and their NNDs"
author: 
  - name: "Jan Linnenbrink"
  - name: "Jakub Nowosad"
date: last-modified
date-format: "[Last Updated:] MMMM, YYYY"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    # page-layout: article
    fig-column: page
    fig-width: 10
    fig-height: 4
    embed-resources: true
knitr:
  opts_chunk:
    warnings: false
execute:
  warning: false
editor: 
  markdown: 
    wrap: sentence
---

```{r}
library(terra)
library(sf)
library(CAST)
library(ggplot2)
library(dplyr)
library(tidyr)
library(cowplot)
library(patchwork)
library(simsam)

get_legend_35 <- function(plot) {
	# return all legend candidates
	legends <- get_plot_component(plot, "guide-box", return_all = TRUE)
	# find non-zero legends
	nonzero <- vapply(legends, \(x) !inherits(x, "zeroGrob"), TRUE)
	idx <- which(nonzero)
	# return first non-zero legend if exists, and otherwise first element (which will be a zeroGrob)
	if (length(idx) > 0) {
		return(legends[[idx[1]]])
	} else {
		return(legends[[1]])
	}
}

theme_set(theme_bw(base_size = 15))
```

## Geodist

```{r}
#| fig-height: 6

k = 5
rast_grid = terra::rast(xmin = 0, xmax = 200, ymin = 0, ymax = 200, ncols = 200, nrows = 200)

study_area <- st_bbox(rast_grid)

ppoints <- spatSample(rast_grid, 1000, "regular", as.points = TRUE) |>
	st_as_sf()

set.seed(100)
sample_r <- sam_field(rast_grid, 100, "random") |>
	mutate(type = "random", cv_folds = caret::createFolds(type, k = k, list = FALSE, returnTrain = FALSE))
sample_c <- sam_field(rast_grid, 100, "clustered", nclusters = 5, value = 20) |>
	mutate(type = "clustered", cv_folds = caret::createFolds(type, k = k, list = FALSE, returnTrain = FALSE))

samples <- list("random" = sample_r, "clustered" = sample_c)

cv_distances <- lapply(names(samples), function(t) {
	s <- samples[[t]]
	cvfolds <- s$cv_folds

	tmp <- list()
	for (i in unique(cvfolds)) {
		tmp[[i]] <- which(cvfolds == i)
	}

	cvfolds <- tmp

	d_cv <- c()
	for (i in 1:length(cvfolds)) {
		d_cv_tmp <- sf::st_distance(s[cvfolds[[i]], ], s[-cvfolds[[i]], ])
		d_cv <- c(d_cv, apply(d_cv_tmp, 1, min))
	}

	dists_cv <- data.frame(dist = d_cv, what = "CV-distances", design = unique(s$type))
}) |>
	do.call(what = "rbind")

# Other distances
sampletosample <- lapply(names(samples), function(t) {
	s <- samples[[t]]
	d <- sf::st_distance(s)
	diag(d) <- Inf
	min_d <- apply(d, 1, min)
	data.frame(dist = min_d, what = factor("sample-to-sample"), design = unique(s$type))
}) |>
	do.call(what = "rbind")

# sample to prediction distance
sampletoprediction <- lapply(names(samples), function(t) {
	s <- samples[[t]]
	d0 <- sf::st_distance(ppoints, s)
	min_d0 <- apply(d0, 1, min)
	data.frame(dist = min_d0, what = factor("prediction-to-sample"), design = unique(s$type))
}) |>
	do.call(what = "rbind")

geodistances <- rbind(sampletosample, sampletoprediction, cv_distances)
rownames(geodistances) <- NULL
geodistances <- mutate(
	geodistances,
	"CV" = ifelse(grepl("CV", what), TRUE, FALSE),
	"design" = factor(design, levels = c("random", "clustered"))
)

# from https://github.com/HannaMeyer/CAST/blob/master/R/plot.R#L456
labs <- c("sample-to-sample", "prediction-to-sample", "CV-distances", "test-to-sample")
myColors <- RColorBrewer::brewer.pal(4, "Dark2")
names(myColors) <- labs
xlabs <- "geographic distances (m)"
geodist_plot <- ggplot(data = geodistances, aes(x = dist, group = what, fill = what)) +
	geom_density(adjust = 1.5, alpha = .5, stat = "density", lwd = 0.3) +
	scale_fill_manual(name = "Prediction distance:", values = myColors) +
	theme_bw() +
	xlab(xlabs) +
	scale_x_log10() +
	facet_wrap(~design, ncol = 1, strip.position = "right") +
	theme(
		legend.position = "bottom",
		legend.box = "vertical",
		legend.spacing.y = unit(2, "mm"),
		legend.title = element_text(size = 15, hjust = 0.5),
		legend.text = element_text(size = 13.5),
		legend.title.position = "top",
		strip.text = element_text(size = 15),
		panel.grid = element_blank()
	)

legend <- get_legend_35(geodist_plot)
geodist_plot <- geodist_plot +
	theme(legend.position = "None")

samples_df <- do.call("rbind", samples) |>
	mutate(type = factor(type, levels = names(samples)))

maps <- ggplot(samples_df) +
	geom_sf(data = st_as_sfc(study_area), colour = "transparent", fill = "transparent") +
	geom_sf() +
	facet_wrap(~type, ncol = 1) +
	theme(panel.grid = element_blank(), strip.text = element_blank())

maps_geodist <- maps + geodist_plot + plot_annotation(tag_levels = 'A') & theme(plot.tag = element_text(size = 16))
combined_plot_legend <- plot_grid(maps_geodist, legend[[1]], ncol = 1, rel_heights = c(1, 0.2))

combined_plot_legend
ggsave("figures/geodist_sim.pdf", combined_plot_legend, width = 8, height = 6)
ggsave("figures/geodist_sim.png", combined_plot_legend, width = 8, height = 6)

```

**Comparison of random and clustered sampling designs and their effects on distance distributions and cross-validation.**
**[Panel a) shows sampling locations for two different designs: random and clustered. Panel b) shows the respective density distributions of nearest-neighbor distances between samples (green), between prediction points and samples (orange) and between cross-validation (CV) folds. For randomly distributed samples, the CV-distances closely resemble the prediction distances. For clustered sampling, however, the CV-distances are much shorter, and thus won't yield reliable estimates of map accuracy.]**

## Spatial cross-validation

```{r}
library(blockCV)
set.seed(2024 - 01 - 14)

spatcluster = cv_cluster(x = sample_r, k = k)
sample_r2 = mutate(sample_r, type = "spatial", cv_folds = spatcluster$folds_id)
sample_r_all = rbind(sample_r, sample_r2)
sample_r_all2 = rename(sample_r_all, fold = cv_folds, split = type)

scv_compare_map = ggplot() +
	geom_sf(data = st_as_sfc(study_area), colour = "transparent", fill = "transparent") +
	geom_sf(data = sample_r_all2, alpha = 0.2, shape = 16) +
	geom_sf(data = sample_r_all, color = "#FF6F61", size = 1.5, alpha = 1) +
	facet_grid(
		type ~ cv_folds,
		# switch = "y",
		labeller = labeller(
			cv_folds = function(x) paste0("Fold: ", x) # Add "Fold: " before each fold
		)
	) +
	ylab("Cross-validation type") +
	scale_y_continuous(sec.axis = dup_axis(labels = NULL)) +
	theme_bw() +
	theme(
		panel.grid = element_blank(),
		strip.text = element_text(size = 14),
		axis.title.y = element_text(size = 16),
		axis.title.y.left = element_blank(),
		axis.ticks.y = element_blank()
	)

scv_compare_map
ggsave("figures/scv_compare_map.pdf", scv_compare_map, width = 8, height = 4)
ggsave("figures/scv_compare_map.png", scv_compare_map, width = 8, height = 4)
```

**Comparison between non-spatial (random) and spatial cross-validation (spatial clustering) methods.**

## Strategies for map accuracy estimation

```{r}
get_random_sample = function(sample_size, r_df) {
	sample_df_random = r_df[sample(nrow(r_df), sample_size), ]
	sample_df_random$sample_size = sample_size
	return(sample_df_random)
}

get_clustered_sample = function(sample_size, r, r_df, type = "clustered") {
	if (type == "clustered") {
		n_clusters = 100
		m = 25
	} else {
		n_clusters = 120
		m = 10
	}

	create_clustered_sample = function(r_df) {
		two_stage = function(sframe, psu, n, m) {
			# function for the two-stage clustered sampling
			units = sample.int(nrow(sframe), size = n, replace = TRUE)
			mypsusample = sframe[units, psu]

			ssunits = unlist(lapply(mypsusample, function(psunit) {
				sample(which(sframe[, psu] == psunit), size = m, replace = FALSE)
			}))

			psudraw = rep(c(1:n), each = m)
			mysample = data.frame(sframe[ssunits, ], psudraw)
			return(mysample)
		}
		r_grid = r_df[c("x", "y")]
		tkmean = kmeans(r_grid, centers = n_clusters, nstart = 5)
		r_grid$psu = tkmean$cluster
		n_psu = ceiling(sample_size / m)
		mysample = two_stage(sframe = r_grid, psu = "psu", n = n_psu, m = m)
		mysample = sf::st_as_sf(mysample, coords = c("x", "y"))
		return(mysample)
	}
	sample_clustered = create_clustered_sample(r_df)
	sample_df_clustered = na.omit(terra::extract(r, sample_clustered, xy = TRUE, ID = FALSE))
	sample_df_clustered$sample_size = sample_size
	return(sample_df_clustered)
}

generate_prob_sampling = function(design) {
	samples_subset = samples |>
		filter(design == {{ design }})

	# Select appropriate raster
	if (design != "extrapolation") {
		sampling_raster = r
	} else {
		sampling_raster = r_east
	}

	# Standard validation sample
	val_SRS = terra::spatSample(
		sampling_raster,
		size = floor(nrow(samples_subset) * (1 - 0.7)),
		method = "random",
		na.rm = TRUE,
		as.points = TRUE
	) |>
		st_as_sf() |>
		mutate(name = "design-based") |>
		select(name)

	# Training sample
	samples_prob = samples_subset |>
		slice_sample(n = floor(nrow(samples_subset) * 0.7)) |>
		mutate(name = "training_prob") |>
		select(name)

	# Combine results
	combined = bind_rows(val_SRS, samples_prob) |>
		mutate(
			design = design
		)
	return(combined)
}
generate_folds = function(design) {
	samples_subset = samples |>
		filter(design == {{ design }})

	# random CV
	random_folds = sample(1:10, replace = TRUE, size = nrow(samples_subset))

	# spatial CV
	mdist = dist(st_coordinates(samples_subset))
	hc = hclust(mdist, method = "complete")
	# val.dist = 350
	# d = val.dist * 1000
	d = 55
	spatial_folds = cutree(hc, h = d)

	# kNNDM CV
	if (design != "extrapolation") {
		sarea = study_area
	} else if (design == "extrapolation") {
		sarea = study_area_east
	}
	knndm_folds = CAST::knndm(samples_subset, modeldomain = sarea)$clusters

	# Combine results
	data.frame(
		design = design,
		randomCV = random_folds,
		spatialCV = spatial_folds,
		knndmCV = knndm_folds
	) |>
		cbind(samples_subset[, "geometry"])
}

r = terra::rast(xmin = 0, xmax = 200, ymin = 0, ymax = 200, ncols = 200, nrows = 200, vals = 1)
r_df = as.data.frame(r, xy = TRUE, na.rm = TRUE)

r_west = terra::crop(r, terra::ext(0, 100, 0, 200))
r_west_df = as.data.frame(r_west, xy = TRUE, na.rm = TRUE)

r_east = terra::crop(r, terra::ext(100, 200, 0, 200))

study_area = st_as_sfc(st_bbox(r))
study_area_east = st_as_sfc(st_bbox(r_east))

study_areas_all = rbind(
	study_area |> st_as_sf() |> mutate(design = "random"),
	study_area |> st_as_sf() |> mutate(design = "biased"),
	study_area |> st_as_sf() |> mutate(design = "clustered"),
	study_area_east |> st_as_sf() |> mutate(design = "extrapolation")
) |>
	mutate(design = factor(design, levels = c("random", "biased", "clustered", "extrapolation")))


sample_size = 300

# regular_sample -----------------------------------------------------------
set.seed(1)
sample_df_random = get_random_sample(sample_size, r_df) |>
	mutate(design = "random")
sample_r = st_as_sf(sample_df_random, coords = c("x", "y"))
# random_sample -----------------------------------------------------------
sample_df_biased = get_clustered_sample(sample_size, r, r_df, type = "biased") |>
	mutate(design = "biased")
sample_b = st_as_sf(sample_df_biased, coords = c("x", "y"))
# clustered_sample --------------------------------------------------------
sample_df_clustered = get_clustered_sample(sample_size, r, r_df, type = "clustered") |>
	mutate(design = "clustered")
sample_c = st_as_sf(sample_df_clustered, coords = c("x", "y"))
# extrapolation sample --------------------------------------------------
sample_df_extra = get_clustered_sample(sample_size, r_west, r_west_df, type = "clustered") |>
	mutate(design = "extrapolation")
sample_e = st_as_sf(sample_df_extra, coords = c("x", "y"))

samples <- list("random" = sample_r, "biased" = sample_b, "clustered" = sample_c, "extrapolation" = sample_e) |>
	do.call(what = "rbind") |>
	mutate(design = factor(design, levels = c("random", "biased", "clustered", "extrapolation")))


set.seed(10)

prob_sampling = lapply(unique(samples$design), generate_prob_sampling) |>
	do.call(what = "rbind") |>
	select(design, cv_name = name, geometry) |>
	mutate(
		value = ifelse(cv_name == "design-based", 1, 2),
		cv_name = ifelse(value == 2, "design-based", cv_name)
	)

folds = lapply(unique(samples$design), generate_folds) |>
	do.call(what = "rbind") |>
	pivot_longer(c(randomCV, spatialCV, knndmCV)) |>
	select(design, cv_name = name, value, geometry) |>
	mutate(
		cv_name = case_when(
			cv_name == "randomCV" ~ "random CV",
			cv_name == "spatialCV" ~ "spatial CV",
			cv_name == "knndmCV" ~ "prediction-domain adaptive CV"
		)
	)

folds = bind_rows(prob_sampling, folds)

folds = folds |>
	mutate(
		cv_name = factor(
			cv_name,
			levels = c(
				"design-based",
				"random CV",
				"spatial CV",
				"prediction-domain adaptive CV"
			)
		)
	) |>
	st_as_sf()


point_cols = Polychrome::alphabet.colors(26)[c(23, 15, 1:14, 16:22, 24:26)]
names(point_cols) = 1:26

validation_folds = ggplot() +
	geom_sf(data = study_areas_all, fill = "#FFD700", aes(colour = "prediction area"), linewidth = 1, alpha = 0.05) +
	scale_colour_manual("prediction area", values = "#FFD700", name = NULL) +
	ggnewscale::new_scale_colour() +
	geom_sf(data = folds, aes(color = as.factor(value)), size = 0.6) +
	scale_colour_manual(values = point_cols, guide = NULL) +
	facet_grid(
		cv_name ~ design,
		labeller = ggplot2::labeller(
			.rows = ggplot2::label_wrap_gen(width = 20)
		)
	) +
	theme_bw(base_size = 12) +
	theme(
		panel.grid = element_blank(),
		axis.text.x = element_text(angle = 45, hjust = 1),
		legend.position = "bottom"
	)

validation_folds
ggsave("figures/validation_folds.pdf", validation_folds, width = 8, height = 8)
ggsave("figures/validation_folds.png", validation_folds, width = 8, height = 8)
```

*Comparison of different strategies for map accuracy estimation: design-based sampling, random cross-validation, spatial cross-validation, and prediction-domain adaptive cross-validation, under a spectrum of sampling designs (from random to extrapolation).*

