---
title: "Figure: CV methods and their NNDs"
author: "Jan Linnenbrink"
date: last-modified
date-format: "[Last Updated:] MMMM, YYYY"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    # page-layout: article
    fig-column: page
    fig-width: 10
    fig-height: 4
    embed-resources: true
knitr:
  opts_chunk:
    warnings: false
execute:
  warning: false
editor: 
  markdown: 
    wrap: sentence
---

```{r}
library(terra)
library(sf)
library(CAST)
library(ggplot2)
library(dplyr)
library(cowplot)
library(patchwork)
library(simsam)

get_legend_35 <- function(plot) {
	# return all legend candidates
	legends <- get_plot_component(plot, "guide-box", return_all = TRUE)
	# find non-zero legends
	nonzero <- vapply(legends, \(x) !inherits(x, "zeroGrob"), TRUE)
	idx <- which(nonzero)
	# return first non-zero legend if exists, and otherwise first element (which will be a zeroGrob)
	if (length(idx) > 0) {
		return(legends[[idx[1]]])
	} else {
		return(legends[[1]])
	}
}

theme_set(theme_bw(base_size = 15))
```

## Geodist

```{r}
#| fig-height: 6

k = 10
rast_grid = terra::rast(xmin = 0, xmax = 200, ymin = 0, ymax = 200, ncols = 200, nrows = 200)

study_area <- st_bbox(rast_grid)

ppoints <- spatSample(rast_grid, 1000, "regular", as.points = TRUE) |>
	st_as_sf()

set.seed(100)
sample_r <- sam_field(rast_grid, 100, "random") |>
	mutate(type = "random", cv_folds = caret::createFolds(type, k = k, list = FALSE, returnTrain = FALSE))
sample_c <- sam_field(rast_grid, 100, "clustered", nclusters = 5, value = 20) |>
	mutate(type = "clustered", cv_folds = caret::createFolds(type, k = k, list = FALSE, returnTrain = FALSE))

samples <- list("random" = sample_r, "clustered" = sample_c)

cv_distances <- lapply(names(samples), function(t) {
	s <- samples[[t]]
	cvfolds <- s$cv_folds

	tmp <- list()
	for (i in unique(cvfolds)) {
		tmp[[i]] <- which(cvfolds == i)
	}

	cvfolds <- tmp

	d_cv <- c()
	for (i in 1:length(cvfolds)) {
		d_cv_tmp <- sf::st_distance(s[cvfolds[[i]], ], s[-cvfolds[[i]], ])
		d_cv <- c(d_cv, apply(d_cv_tmp, 1, min))
	}

	dists_cv <- data.frame(dist = d_cv, what = "CV-distances", design = unique(s$type))
}) |>
	do.call(what = "rbind")

# Other distances
sampletosample <- lapply(names(samples), function(t) {
	s <- samples[[t]]
	d <- sf::st_distance(s)
	diag(d) <- Inf
	min_d <- apply(d, 1, min)
	data.frame(dist = min_d, what = factor("sample-to-sample"), design = unique(s$type))
}) |>
	do.call(what = "rbind")

# sample to prediction distance
sampletoprediction <- lapply(names(samples), function(t) {
	s <- samples[[t]]
	d0 <- sf::st_distance(ppoints, s)
	min_d0 <- apply(d0, 1, min)
	data.frame(dist = min_d0, what = factor("prediction-to-sample"), design = unique(s$type))
}) |>
	do.call(what = "rbind")

geodistances <- rbind(sampletosample, sampletoprediction, cv_distances)
rownames(geodistances) <- NULL
geodistances <- mutate(
	geodistances,
	"CV" = ifelse(grepl("CV", what), TRUE, FALSE),
	"design" = factor(design, levels = c("random", "clustered"))
)

# from https://github.com/HannaMeyer/CAST/blob/master/R/plot.R#L456
labs <- c("sample-to-sample", "prediction-to-sample", "CV-distances", "test-to-sample")
myColors <- RColorBrewer::brewer.pal(4, "Dark2")
names(myColors) <- labs
xlabs <- "geographic distances (m)"
geodist_plot <- ggplot(data = geodistances, aes(x = dist, group = what, fill = what)) +
	geom_density(adjust = 1.5, alpha = .5, stat = "density", lwd = 0.3) +
	scale_fill_manual(name = "distance function", values = myColors) +
	theme_bw() +
	xlab(xlabs) +
	scale_x_log10() +
	facet_wrap(~design, ncol = 1, strip.position = "right") +
	theme(
		legend.position = "bottom",
		legend.box = "vertical",
		legend.spacing.y = unit(2, "mm"),
		legend.title = element_text(size = 15, hjust = 0.5),
		legend.text = element_text(size = 13.5),
		legend.title.position = "top",
		strip.text = element_text(size = 15),
		panel.grid = element_blank()
	)

legend <- get_legend_35(geodist_plot)
geodist_plot <- geodist_plot +
	theme(legend.position = "None")

samples_df <- do.call("rbind", samples) |>
	mutate(type = factor(type, levels = names(samples)))

maps <- ggplot(samples_df) +
	geom_sf(data = st_as_sfc(study_area), colour = "black", fill = "transparent") +
	geom_sf() +
	facet_wrap(~type, ncol = 1) +
	theme(panel.grid = element_blank(), strip.text = element_blank())

maps_geodist <- maps + geodist_plot + plot_annotation(tag_levels = 'A') & theme(plot.tag = element_text(size = 16))
combined_plot_legend <- plot_grid(maps_geodist, legend[[1]], ncol = 1, rel_heights = c(1, 0.2))

combined_plot_legend
ggsave("figures/geodist_sim.pdf", combined_plot_legend, width = 8, height = 6)
ggsave("figures/geodist_sim.png", combined_plot_legend, width = 8, height = 6)

```

**Comparison of random and clustered sampling designs and their effects on distance distributions and cross-validation.**
**[Panel a) shows sampling locations for two different designs: random and clustered. Panel b) shows the respective density distributions of nearest-neighbor distances between samples (green), between prediction points and samples (orange) and between cross-validation (CV) folds. For randomly distributed samples, the CV-distances closely resemble the prediction distances. For clustered sampling, however, the CV-distances are much shorter, and thus won't yield reliable estimates of map accuracy.]**