---
title: "Extrapolation vs Interpolation: A continuoum"
author: 
  - name: "Jan Linnenbrink"
date: last-modified
date-format: "[Last Updated:] MMMM, YYYY"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    # page-layout: article
    fig-column: page
    fig-width: 10
    fig-height: 4
    embed-resources: true
knitr:
  opts_chunk:
    warnings: false
execute:
  warning: false
editor: 
  markdown: 
    wrap: sentence
---

```{r}
library(terra)
library(sf)
library(CAST)
library(ggplot2)
library(dplyr)
library(cowplot)
library(patchwork)
library(simsam)

get_legend_35 <- function(plot) {
	# return all legend candidates
	legends <- get_plot_component(plot, "guide-box", return_all = TRUE)
	# find non-zero legends
	nonzero <- vapply(legends, \(x) !inherits(x, "zeroGrob"), TRUE)
	idx <- which(nonzero)
	# return first non-zero legend if exists, and otherwise first element (which will be a zeroGrob)
	if (length(idx) > 0) {
		return(legends[[idx[1]]])
	} else {
		return(legends[[1]])
	}
}

theme_set(theme_bw(base_size = 15))
```


```{r}
plot.knndm.integral <- function(Gij, Gj, ...) {
	# Prepare data for plotting: Gij function
	Gij_df <- data.frame(r = Gij[order(Gij)])
	Gij_df$Function <- "1_Gij(r)"

	# Prepare data for plotting: G function
	Gj_df <- data.frame(r = Gj[order(Gj)])
	Gj_df$Function <- "2_Gj(r)"

	# Merge data for plotting
	Gplot <- rbind(Gij_df, Gj_df)

	n = length(Gij_df$r)
	dfx = data.frame(sort(Gij_df$r), cumsum(rep(1 / n, n)), "x")
	n1 = length(Gj_df$r)
	dfy = data.frame(sort(Gj_df$r), cumsum(rep(1 / n1, n1)), "y")
	colnames(dfx) = colnames(dfy) = c("point", "ecdf", "sample")
	df = rbind(dfy, dfx)

	combined = sort(c(Gij_df$r, Gj_df$r))

	midpoints = combined[-(2 * n)] + diff(combined) / 2
	fhat = ehat = dhat = width = numeric(2 * n - 1)
	for (i in 1:(2 * n - 1)) {
		point = midpoints[i]
		width[i] = (point - combined[i]) * 2
		fhat[i] = mean(Gj_df$r < point)
		ehat[i] = mean(Gij_df$r < point)
		dhat[i] = mean(combined < point)
	}
	gap = fhat - ehat

	x1 = midpoints - 0.5 * width
	x2 = midpoints + 0.5 * width
	y1 = ehat
	y2 = fhat
	block = 1:(2 * n - 1)
	alpha = 1 / (dhat * (1 - dhat))

	df3.1 = cbind(x1, y1, block, alpha)
	df3.2 = cbind(x1, y2, block, alpha)
	df3.3 = cbind(x2, y2, block, alpha)
	df3.4 = cbind(x2, y1, block, alpha)

	df3 = rbind(df3.1, df3.2, df3.3, df3.4)
	colnames(df3) = c("point", "ecdf", "block", "weights")
	df3 = as_tibble(df3) |> arrange(block)

	# Plot
	ggplot2::ggplot(data = Gplot) +
		ggplot2::geom_vline(xintercept = 0, lwd = 0.1, ggplot2::aes_string(x = "r", group = "Function", col = "Function")) +
		ggplot2::geom_hline(yintercept = 0, lwd = 0.1, ggplot2::aes_string(x = "r", group = "Function", col = "Function")) +
		ggplot2::geom_hline(yintercept = 1, lwd = 0.1, ggplot2::aes_string(x = "r", group = "Function", col = "Function")) +
		geom_polygon(data = df3, aes(x = point, y = ecdf, group = block), alpha = 0.7, fill = "orange") +
		ggplot2::stat_ecdf(geom = "step", lwd = 1, ggplot2::aes_string(x = "r", group = "Function", col = "Function")) +
		ggplot2::scale_colour_manual(
			name = "geographic distance:",
			values = c("#000000", "#56B4E9"),
			labels = c("prediction-to-sample", "sample-to-sample")
		) +
		ggplot2::ylab(expression(paste(hat(G)[ij](r), ", ", hat(G)[j]^"*" * "(r,L)", ", ", hat(G)[j](r)))) +
		ggplot2::theme(panel.grid = element_blank())
}
```

## Extrapolation vs Interpolation: A continuoum rather than a binary classification

```{r}
#| fig-height: 6

set.seed(10)

# Simulate samples
rast_grid <- terra::rast(xmin = 0, xmax = 200, ymin = 0, ymax = 200, ncols = 200, nrows = 200)
rast_grid_transfer <- terra::rast(xmin = 200, xmax = 400, ymin = 0, ymax = 200, ncols = 200, nrows = 200)

study_area <- st_bbox(rast_grid)
study_area_transfer <- st_bbox(rast_grid_transfer)

ppoints_inter <- spatSample(rast_grid, 1000, "regular", as.points = TRUE) |>
	st_as_sf() |>
	mutate(type = "interpolation")
ppoints_transfer <- spatSample(rast_grid_transfer, 1000, "regular", as.points = TRUE) |>
	st_as_sf() |>
	mutate(type = "extrapolation")

ppoints <- list("inter" = ppoints_inter, "extra" = ppoints_transfer)

length_out <- 5
clustering_min <- 20
clustering_max <- 100

clustering <- as.integer(round(exp(seq(log(clustering_min), log(clustering_max), length.out = length_out))))[1:4]

# clustering <- seq(10, 100, 10)[2:10]
samples <- lapply(clustering, function(x) {
	sam_field(rast_grid, size = 100, type = "clustered", value = x, nclusters = 4) |>
		mutate(clustering = x)
}) |>
	do.call(what = "rbind")


# calculate NNDs between samples and between prediction and samples
geodistances <- lapply(clustering, function(cl) {
	s <- samples |>
		filter(clustering == cl)

	# sample to sample NND
	d <- sf::st_distance(s)
	diag(d) <- Inf
	min_d <- apply(d, 1, min)
	sampletosample <- data.frame(dist = min_d, what = factor("sample-to-sample"), clustering = unique(s$clustering))

	# prediction to sample NND
	d0 <- sf::st_distance(ppoints$inter, s)
	min_d0 <- apply(d0, 1, min)

	sampletoprediction <- data.frame(
		dist = min_d0,
		what = factor("prediction-to-sample"),
		clustering = unique(s$clustering)
	)

	geod_df <- rbind(sampletosample, sampletoprediction)
	rownames(geod_df) <- NULL

	# add transfer for the most clustered design
	if (cl == clustering[[1]]) {
		# sample to sample NND
		d <- sf::st_distance(s)
		diag(d) <- Inf
		min_d <- apply(d, 1, min)
		sampletosample <- data.frame(dist = min_d, what = factor("sample-to-sample"), clustering = 0)

		# prediction to sample NND
		d0 <- sf::st_distance(ppoints$extra, s)
		min_d0 <- apply(d0, 1, min)

		sampletoprediction <- data.frame(
			dist = min_d0,
			what = factor("prediction-to-sample"),
			clustering = 0
		)

		geod_df_transfer <- rbind(sampletosample, sampletoprediction)

		geod_df_final <- rbind(geod_df_transfer, geod_df)
		rownames(geod_df_final) <- NULL
	} else {
		geod_df_final <- geod_df
	}

	geod_df_final
}) |>
	do.call(what = "rbind")

# Calculate Wasserstein statistic as measure of extrapolation
W_df <- lapply(c(0, clustering), function(cl) {
	g <- geodistances |> filter(clustering == cl)
	Gij <- g |> filter(what == "prediction-to-sample") |> pull(dist)
	Gj <- g |> filter(what == "sample-to-sample") |> pull(dist)
	W_stat <- twosamples::wass_stat(Gj, Gij)
	data.frame(W = W_stat, clustering = cl)
}) |>
	do.call(what = "rbind")

# Geodistance plots
labs <- c("sample-to-sample", "prediction-to-sample")
myColors <- RColorBrewer::brewer.pal(4, "Dark2")
names(myColors) <- labs
xlabs <- "geographic distances (m)"
geodist_plot_list <- lapply(c(0, clustering), function(cl) {
	Gj <- geodistances |> filter(clustering == cl & what == "sample-to-sample") |> pull(dist)
	Gij <- geodistances |> filter(clustering == cl & what == "prediction-to-sample") |> pull(dist)

	plot.knndm.integral(Gj = Gj, Gij = Gij) +
		xlab("geographic distance") +
		ylab("ECDF") +
		theme(legend.position = "bottom", axis.text = element_text(size = 8)) +
		geom_text(data = W_df |> filter(clustering == cl), aes(x = 200, y = 0.25, label = paste0("W = ", round(W)))) +
		scale_x_continuous(limits = c(0, 300))
})

geodist_legend <- get_legend_35(geodist_plot_list[[1]])
geodist_plot_list_clean <- lapply(geodist_plot_list, function(x) {
	x + theme(legend.position = "None", axis.title = element_blank(), aspect.ratio = 1.2)
})


# Corresponding maps of samples
samples_sf <- do.call("rbind", samples)
study_area_sf <- st_as_sfc(study_area)
study_area_transfer_sf <- st_as_sfc(study_area_transfer)

samples_map_list <- lapply(clustering, function(cl) {
	ggplot() +
		geom_sf(data = study_area_sf, fill = "transparent", colour = "transparent") +
		geom_sf(data = samples |> filter(clustering == cl), size = 0.5) +
		facet_wrap(~clustering, nrow = 1) +
		theme(
			panel.grid = element_blank(),
			axis.ticks = element_blank(),
			axis.text = element_blank(),
			strip.text = element_blank()
		)
})


samples_map_transfer <- ggplot() +
	geom_sf(data = samples |> filter(clustering == 20), size = 0.5) +
	geom_sf(data = study_area_sf, fill = "transparent", colour = "transparent") +
	geom_sf(data = study_area_transfer_sf, fill = "transparent", colour = "orange", linewidth = 1) +
	facet_wrap(~"transfer", nrow = 1) +
	theme(
		panel.grid = element_blank(),
		axis.ticks = element_blank(),
		axis.text = element_blank(),
		strip.text = element_blank()
	)

geodist_map_transfer <- wrap_plots(samples_map_transfer, geodist_plot_list_clean[[1]], ncol = 1)
geodist_maps_other <- lapply(seq_along(samples_map_list), function(i) {
	wrap_plots(samples_map_list[[i]], geodist_plot_list_clean[[i + 1]], ncol = 1)
})

geodist_maps_all <- plot_grid(
	plotlist = c(list(geodist_map_transfer), geodist_maps_other),
	nrow = 1,
	rel_heights = c(1, 0.2)
)

geodist_map_final <- plot_grid(geodist_maps_all, geodist_legend, ncol = 1, rel_heights = c(1, 0.1))
geodist_map_final
ggsave("figures/extrapolation_continuum.pdf", geodist_map_final, height = 4, width = 10)

```

**Understanding Extrapolation as an continuum**
**[The upper row shows sampling designs ranging from a setting where transfer to a new region is required (left) over clustered samples towards randomly distributed samples. The lower row shows the difference between the empirical cumulative density functions calculated for the distance between samples (blue, corresponds to interpolation situation) and the ECDF between prediction points and samples (black, the situation the model is applied to). A continuum from strong extrapolation situations on the left (denoted by large integrals between the two ECDFs, orange area and W stat measuring the integral) towards low extrapolation on the right can be observed.]**

